---
title: "Statistical analysis for the geometry of thought - revised"
author: "Clara Wakonigg"
format:
  html:
    embed-resources: true
    code-fold: false
    number-sections: true
toc: true
toc-title: Table of contents
toc-depth: 4
toc-location: left
execute: 
  eval: false
  echo: true
output:
  html_document:
  output-file: stats_analysis_s1_complete.html
---
```{r}
# ---- Packages ----
# Data wrangling + plotting
library(tidyverse)

# Mixed models
library(lme4)
library(glmmTMB)
library(modelbased)

# Model summaries / tidying / prediction
library(broom)
library(broom.mixed)

# Distances / geometry / clustering / visualisation (load only if used)
library(pracma)        # geo_median
library(cluster)       # pam() 
library(fpc)           # cluster stability

# Plot extras
library(ggside)

# cat PCA
library(Gifi)

# ICC
library(performance)
```

# Load the data

```{r}
data_path <- file.path(dirname(getwd()), "data", "master.csv")

if (!file.exists(data_path)) {
  stop("Dataset is currently not publicaly available.")
}

esm_traits <- read.csv(data_path) |>
  dplyr::mutate(ID = forcats::as_factor(ID), 
                Stimulus_Type = forcats::fct_relevel(Stimulus_Type, "Baseline", "pos", "neg"), 
                GENDER = ifelse(GENDER == 1,"Female", "Male"))
```

```{r}
cesd_by_id <- esm_traits |>
  dplyr::select(ID, CESD_stnd, CESD, AGE, GENDER) |>
  distinct(ID, .keep_all = TRUE) |>
  mutate(AGE_stnd = scale(AGE))
```

```{r}
# create path string to save results to 
results_path <- file.path(dirname(getwd()), "results")
```

## Standardise 

```{r}
component_cols <- c("Component_1", "Component_2", "Component_3", "Component_4", "Component_5")
esm_traits_z <- esm_traits |>
  mutate(across(
    all_of(component_cols),
    ~ (.x - mean(.x, na.rm = TRUE)) / sd(.x, na.rm = TRUE),
    .names = "{.col}_z"
  ),
  AGE_stnd = scale(AGE)) 
```

```{r}
# Filter Condition to only include everyday life (session 3) data
esm_traits_base <- esm_traits_z |> 
  dplyr::filter(Stimulus_Type == "Baseline") |>
  dplyr::mutate(ID = droplevels(ID),
                AGE_stnd = scale(AGE))
```

# Descriptive stats

```{r}
# Count number of participants
filtered_id <- esm_traits |> 
  dplyr::group_by(ID) |>
  dplyr::summarise(count = n())
```

```{r}
# Count number of participants by Gender group
esm_traits |> 
  dplyr::distinct(ID, GENDER) |>  
  dplyr::count(GENDER) 
```

```{r}
# Responses per day
response_rate <- esm_traits_base |>
  mutate(
    Date = ymd_hms(Date),  
    Day      = as.Date(Date),          
    Time     = format(Date, "%H:%M:%S")) |>
  group_by(ID, Day) |>
  summarise(n_entries = n(), .groups = "drop")
```

```{r}
# Response rate per day
rr_stats <- response_rate |>
  group_by(ID) |>
  summarise(mean_per_day = mean(n_entries)) 

mean(rr_stats$mean_per_day)
sd(rr_stats$mean_per_day)
range(response_rate$n_entries)
```

```{r}
# How many days with responses
rr_days <- response_rate |>
  group_by(ID) |>
  summarise(n_days = n_distinct(Day), .groups = "drop")

mean(rr_days$n_days)
sd(rr_days$n_days)
range(rr_days$n_days)
```

```{r}
# Proportion of participants with at least 6 days of responses
prop_6days <- rr_days |>
  summarise(
    total_participants = n(),
    at_least_6_days = sum(n_days >= 6),
    proportion = at_least_6_days / total_participants
  )
```

```{r}
# Overall response rate
rr_total <- esm_traits_base |>
  group_by(ID) |>
  summarise(total_responses = n(), .groups = "drop")

mean(rr_total$total_responses)
sd(rr_total$total_responses)
range(rr_total$total_responses)
```

```{r}
# Group response rate
nrow(esm_traits_base) / (n_distinct(esm_traits_base$ID)*42)
```

```{r}
# Distribution of depressive scores
cesd_dis <- cesd_by_id |>
  datawizard::describe_distribution(CESD)

cesd_dis
```
```{r}
# Demographics table
traits_summary <- esm_traits_base |>
  dplyr::distinct(ID, .keep_all = TRUE) |>
  dplyr::select(AGE, GENDER, CESD) |>
  gtsummary::tbl_summary(
    statistic = list(
      gtsummary::all_continuous() ~ "{mean} ({sd})",
      gtsummary::all_categorical() ~ "{n} / {N} ({p}%)"
    ),
    digits = gtsummary::all_continuous() ~ 2
  ) 
traits_summary
```


```{r}
# Demographics table split by dataset
demo <- esm_traits |>
  mutate(block = if_else(Stimulus_Type == "Baseline", "daily", "lab")) |>
  count(Dataset, ID, block) |>
  pivot_wider(names_from = block, values_from = n, values_fill = 0) |>
  left_join(
    esm_traits |>
      group_by(Dataset, ID) |>
      summarise(AGE = first(AGE), GENDER = first(GENDER), CESD = first(CESD), .groups = "drop"),
    by = c("Dataset", "ID")
  )
```

```{r}
fmt <- function(x) sprintf("%.1f (%.1f)", mean(x, na.rm = TRUE), sd(x, na.rm = TRUE))

demo_table <- demo |>
  group_by(Dataset) |>
  summarise(
    `N included`            = n(),
    `Age M (SD)`            = fmt(AGE),
    Gender                  = paste0(sum(GENDER == "Female"), "F / ", sum(GENDER == "Male"), "M"),
    `CES-D M (SD)`          = fmt(CESD),
    `Daily-life obs M (SD)` = fmt(daily),
    `Lab obs M (SD)`        = fmt(lab),
    .groups = "drop"
  )

total_row <- demo |>
  summarise(
    Dataset                 = "Total",
    `N included`            = n(),
    `Age M (SD)`            = fmt(AGE),
    Gender                  = paste0(sum(GENDER == "Female"), "F / ", sum(GENDER == "Male"), "M"),
    `CES-D M (SD)`          = fmt(CESD),
    `Daily-life obs M (SD)` = fmt(daily),
    `Lab obs M (SD)`        = fmt(lab)
  )

bind_rows(demo_table, total_row) |> knitr::kable()
```


```{r}
# Test if demographics differ by dataset
summary(aov(CESD ~ Dataset, data = demo))
summary(aov(AGE ~ Dataset, data = demo))
summary(aov(daily ~ Dataset, data = demo))
chisq.test(table(demo$Dataset, demo$GENDER), simulate.p.value = TRUE)
```

```{r}
# Calculate the median of Negative Intrusive thought
median_c1 <- median(esm_traits_base$Component_1_z)
```

```{r}
# Median split, to be able to count frequency of high and low negative intrusive thought experienced
esm_traits_base <- esm_traits_base |>
  mutate(C1_prop = case_when(
    Component_1_z >= median_c1 ~ "High",
    TRUE ~ "Low"
  ))
```

```{r}
# Categorise depression into Yes or No based on questionnaire threshold
esm_traits_base <- esm_traits_base |>
  mutate(CESD_th = ifelse(CESD >= 16, "Yes", "No"))
```

```{r}
# Calculate proportion of High and Low Negative Intrusive Thought
participant_counts <- esm_traits_base |>
  group_by(ID, C1_prop, CESD_th) |>
  summarise(n = n(), .groups = "drop") |>
  complete(ID, C1_prop, fill = list(n = 0)) |>
  group_by(ID) |>
  tidyr::fill(CESD_th, .direction = "downup") |>
  mutate(prop = n / sum(n)) |>
  ungroup() 
```

```{r}
participant_counts_only <- participant_counts |>
  left_join(cesd_by_id, by = "ID")|>
  mutate(CESD_th = forcats::fct_relevel(CESD_th, "No", "Yes"),
         C1_prop = forcats::fct_relevel(C1_prop, "Low", "High"),
         ID = forcats::as_factor(ID) |> forcats::fct_drop())
```

```{r}
# Count unique participants per group
counts_wide <- participant_counts_only |>
  dplyr::distinct(ID, CESD_th) |>
  dplyr::count(CESD_th) |>
  tidyr::pivot_wider(
    names_from  = CESD_th,
    values_from = n,
    names_prefix = "n_"
  ) |>
  mutate(n_Overall = n_No + n_Yes)

counts_wide
```

```{r}
# Proportion of high and low negative intrusive thoughts in depression groups. 
freq_summary <- esm_traits_base |>
  group_by(ID) |>
  gtsummary::tbl_summary(by = "CESD_th", include = c(CESD, C1_prop, Component_1),
    statistic = list(
      gtsummary::all_continuous() ~ "{mean} ({sd})",
      gtsummary::all_categorical() ~ "{p}"
    ),
    digits = gtsummary::all_continuous() ~ 2,
    missing = "no"
  ) |> gtsummary::add_overall() |>
  gtsummary::modify_spanning_header(c("stat_1", "stat_2") ~ "**Depression Group**") 

freq_summary
```

## Cronback's alpha

```{r}
traits <- read.csv(file.path(dirname(getwd()),"data", "Combined_traits.csv"))|> 
  dplyr::mutate(ID = forcats::as_factor(ID))
```

```{r}
traits <- traits |>
  drop_na(c(24:63))
a_CESD<- ltm::cronbach.alpha(traits[,24:43],standardized = TRUE, CI=T)
CI_CESD <- a_CESD$ci
```

```{r}
# Create the table
alpha_tbl <- data.frame(
  Questionnaire = c("CESD"),
  'Number of Items' = c(a_CESD$p),
  Alpha = c(a_CESD$alpha),
  'Lower CI' = c(CI_CESD[1]), 
  'Upper CI'=c(CI_CESD[2])) |> rempsyc::nice_table()
```

# Emotional Manipulation check

```{r}
affective_summary <- esm_traits_z |>
group_by(Stimulus_Type) |>
  datawizard::describe_distribution(Positive)

affective_test <- lme4::lmer(Positive_stnd ~ Stimulus_Type + (Stimulus_Type|ID), data = esm_traits_z)

parameters::parameters(affective_test)
```
```{r}
affective_test_dep <- lme4::lmer(Positive_stnd ~ Stimulus_Type*CESD_stnd + (Stimulus_Type|ID), data = esm_traits_z)

parameters::parameters(affective_test_dep)
```
# Establish default thought patterns: calculate for each participant the geometric median of their observations

```{r}
centroid_summary <- esm_traits_z |>
  filter(Stimulus_Type == "Baseline") |>
  dplyr::select(ID, Component_1_z:Component_5_z) |>
  group_by(ID) |>
  group_split() |>
  map_dfr(function(df) {
    comp_matrix <- df |>
      dplyr::select(Component_1_z:Component_5_z) |>
      drop_na() |>
      as.matrix()
    geom_vec   <- geo_median(comp_matrix)$p  
    tibble(
      ID = unique(df$ID),
      !!!set_names(as.list(geom_vec),   paste0("Geom_",   1:5))
    )
  })
```

```{r}
# Join the geometric medians to the esm data
esm_centroids <- esm_traits_z |>
  left_join(centroid_summary, by = "ID")
```

```{r}
# Calculate the distance for each observation to the geometric median
esm_centroids <- esm_centroids |>
  rowwise() |>
  mutate(
    Radius_geom = {
      comps <- c_across(Component_1_z:Component_5_z)
      geom_cent <- c_across(Geom_1:Geom_5)
      sqrt(sum((comps - geom_cent)^2))
    }
  ) |>
  ungroup() |>
  mutate(AGE_stnd = scale(AGE), 
         GENDER = as_factor(GENDER))
```

# Similarity analysis

```{r}
participant_ids <- unique(esm_traits_base$ID)
```

```{r}
# Create an empty data frame to store results
within_similarity_df <- data.frame(ID = character(), within_similarity_z = numeric(), stringsAsFactors = FALSE)
```

```{r}
# Loop over each participant ID
for (id in participant_ids) {
  # Subset data for the current participant
  participant_data <- esm_traits_base |>
    dplyr::filter(ID == id) |>
    dplyr::select(c("Component_1_z":"Component_5_z"))
  # Convert to matrix
  participant_matrix <- as.matrix(participant_data)
  # Only proceed if participant has more than 1 measurement
  if (nrow(participant_matrix) > 1) {
    # Compute trial × trial correlation
    trial_cor <- cor(t(participant_matrix))
    # Extract lower triangle
    lower_vals <- trial_cor[lower.tri(trial_cor)]
    #Fisher transform correlations before averaging
    lower_vals_clipped <- pmin(pmax(lower_vals, -0.9999), 0.9999)
    z_vals <- atanh(lower_vals_clipped)
    # Compute mean z-value
    mean_z <- mean(z_vals, na.rm = TRUE)
    # Add result to the data frame
    within_similarity_df <- rbind(within_similarity_df, data.frame(ID = id, within_similarity_z = mean_z))
  }
}
mean_within_similarity <- mean(within_similarity_df$within_similarity_z)
```

```{r}
# Between subject similarity
mean_vectors <- esm_traits_base |>
  group_by(ID) |>
  summarise(across(c("Component_1_z":"Component_5_z"), \(x) mean(x, na.rm = TRUE))) |>
  ungroup()

mean_matrix <- as.matrix(mean_vectors[, -1])
```

```{r}
# Correlation matrix 
between_cor <- cor(t(mean_matrix))
```

```{r}
# extract lower triangle
lower_between <- between_cor[lower.tri(between_cor)]
```

```{r}
# Fisher z
z_between <- atanh(lower_between)
mean_between_similarity <- mean(z_between, na.rm = TRUE)
within_z <- within_similarity_df$within_similarity_z
between_z <- z_between  # the full lower triangle of between-subject correlations, Fisher-z transformed
```

```{r}
# Compute the observed difference
observed_diff <- mean(within_z) - mean(between_z)
```

```{r}
# Combine all values into one vector
all_z_values <- c(within_z, between_z)
group_labels <- c(rep("within", length(within_z)), rep("between", length(between_z)))
```

```{r}
# Set number of permutations
n_perm <- 10000
perm_diffs <- numeric(n_perm)

set.seed(123)

for (i in 1:n_perm) {
  shuffled_labels <- sample(group_labels)
  group1 <- all_z_values[shuffled_labels == "within"]
  group2 <- all_z_values[shuffled_labels == "between"]
  perm_diffs[i] <- mean(group1) - mean(group2)
}
```

```{r}
# p-value: how often is permuted diff ≥ observed diff?
p_value <- mean(perm_diffs >= observed_diff)
perm_df <- data.frame(perm_diff = perm_diffs)
```
```{r}
mean(within_z)
mean(between_z)
observed_diff
p_value
```

```{r}
# Make the plot
permutation_plot <- ggplot(perm_df, aes(x = perm_diff)) +
  geom_histogram(bins = 50, fill = "grey80", color = "black") +
  geom_vline(xintercept = observed_diff, color = "#990000", linewidth = 1.2) +
  annotate("text",
           x = Inf,
           y = Inf,
           label = "Observed\nDifference",
           hjust = 0.99, vjust = 1, color = "#990000", size = 5) +
  scale_x_continuous(limits = range(c(perm_diffs, observed_diff))) +
  scale_y_continuous(limits = range(0,1000)) +
  labs(
    x = "Mean thought similarity difference (within - between)",
    y = "Frequency"
  ) +
  theme_minimal(base_size = 20) +
  theme(
    plot.margin = unit(c(1, 2, 1, 2), "cm"),
    panel.grid.major     = element_blank(),
    panel.grid.minor     = element_blank(),
    axis.line            = element_line(color = "black"),
    axis.ticks           = element_line(color = "black"),
    # inside top-right
    legend.position      = c(0.997, 0.998),
    legend.justification = c(1, 1),
    legend.key.size      = unit(0.4, "cm"),
    legend.text          = element_text(size = 18),
    legend.spacing.y = unit(0,"cm"),
  ) 
 ggsave(
   filename = file.path(results_path, "permutation_plot_new.svg"),
   plot     = permutation_plot,
   width    = 200,       # cm (1/3 of 84.1 cm)
   height   = 150,     # cm (aspect ratio 10:6)
   units    = "mm",
   dpi      = 300
 )
```

# Baseline models

```{r}
## Negative Intrusive thought modeled from depression levels
dep_c1_lm <- lme4::lmer(Component_1_z ~ CESD_stnd + AGE_stnd + GENDER + (1|ID), data = esm_traits_base)
summary(dep_c1_lm)
parameters::parameters(dep_c1_lm)|> knitr::kable(digits = 3)

anova_c1<- car::Anova(dep_c1_lm, type = 3, test.statistic = "F") 
anova_tbl <- anova_c1 |> 
  as.data.frame() |>
  rownames_to_column(var = "term")

effectsize::eta_squared(dep_c1_lm, partial = TRUE, alternative ="two.sided")
anova_c1
```

```{r}
# Plot the relationship
depression_model_plot <- ggplot(
  esm_traits_base,
  aes(x = CESD_stnd, y = Component_1_z)
) +
  geom_point(
    aes(color = "Data point"),
    alpha = 0.15,
    size = 1.5,
    position = position_jitter(
    width  = 0.05,   # small jitter on CESD axis
    height = 0.00,   # small jitter on Component axis
    seed = 123
  )
  ) +
  geom_smooth(
    aes(color = "Best fit", fill = "95% CI"),
    method = "lm",
    linewidth = 1,
    se = TRUE,
    formula = y ~ x,
  ) +
  # ---- SIDE DENSITIES (SAFE) ----
  geom_xsidedensity(
    aes(y = after_stat(density)),
    fill = "grey70",
    alpha = 0.4
  ) +
  geom_ysidedensity(
    aes(x = after_stat(density)),
    fill = "grey70",
    alpha = 0.4
  ) +
  scale_color_manual(
    name = NULL,
    values = c(
      "Data point" = "black",
      "Best fit"   = "#990000"
    )
  ) +
  scale_fill_manual(
    name = NULL,
    values = c("95% CI" = "#e06666"),
    guide = guide_legend(override.aes = list(color = NA))
  ) +
  labs(
    x = "Depression score (CESD)",
    y = "Negative Intrusive Thought (Component 1)"
  ) +
  theme_minimal(base_size = 20) +
  theme(
    ggside.panel.scale = 0.25,
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.line = element_line(color = "black"),
    axis.ticks = element_line(color = "black"),
    legend.position = c(0.997, 0.998),
    legend.justification = c(1, 1)
  )
depression_model_plot
```

```{r}
ggsave(
  filename = file.path(results_path, "depression_c1_scatter.svg"),
  plot     = depression_model_plot,
  width    = 280,       
  height   = 150,     
  units    = "mm",
  dpi      = 300
)
```

# Frequency Models

```{r}
esm_binomial <- esm_traits_base |>
  group_by(ID) |>
  summarise(
    high = sum(C1_prop == "High", na.rm = TRUE),
    total = n(),
    CESD_stnd = first(CESD_stnd),
    AGE = first(AGE),
    GENDER = first(GENDER)
  ) |>
  mutate(low = total - high,
         AGE_stnd = scale(AGE))
```

### Model

```{r}
model_binom <- glm(cbind(high, total - high) ~ CESD_stnd + AGE_stnd + GENDER ,
                   family = quasibinomial(link = "logit"),
                   data = esm_binomial)
summary(model_binom)
or_bi <- parameters::parameters(model_binom, exponentiate = TRUE) 
or_bi |> knitr::kable(digits = 3)
```

```{r}
binomial_plot <-ggplot(esm_binomial, aes(x = CESD_stnd, y = high / total)) +
  geom_point(color = "black", alpha = 0.6, size = 1.5) +
  stat_smooth(method = "glm", 
              method.args = list(family = "quasibinomial"), 
              se = TRUE, 
              color = "#990000") +
  labs(
    y = "Proportion of High Negative Intrusive Thoughts", 
    x = "Depression Score (CESD)"
  ) +
  theme_minimal(base_size = 18) 

binomial_plot
```

```{r}
ggsave(
  filename = file.path(results_path, "binomial_plot_c1.jpg"),
  plot     = binomial_plot,
  width    = 240,       # cm (1/3 of 84.1 cm)
  height   = 150,     # cm (aspect ratio 10:6)
  units    = "mm",
  dpi      = 300
)
```

# Cluster analysis

```{r}
# Cluster analysis of the centroid coordinates of participants
centroid_dep <- centroid_summary |>
  left_join(esm_traits_z |> dplyr::select(ID, AGE, GENDER, CESD, CESD_stnd, Dataset) |>
              distinct(), by = "ID")

sil_results <- data.frame(k = 2:8, avg_sil = NA)

for (k in 2:8) {
  pam_fit <- pam(centroid_dep[, c("Geom_1","Geom_2","Geom_3","Geom_4","Geom_5")],
                 k = k, metric = "euclidean")
  sil_obj <- silhouette(pam_fit)
  sil_results$avg_sil[sil_results$k == k] <- mean(sil_obj[, "sil_width"])
}

sil_results

plot(sil_results$k, sil_results$avg_sil, type = "b",
     xlab = "k", ylab = "Average silhouette width")
```

```{r}
# Plot the silhouette width to identify best K cluster
sil_plot <- ggplot(sil_results, aes(x = k, y = avg_sil)) +
  geom_line(linewidth = 0.6, linetype = 1) +   # line first, underneath
  geom_point(shape = 21, size = 3, fill = "white", stroke = 1) +  # dots above
  labs(x = "k", y = "Average silhouette width") +
  theme_minimal(base_size = 18)

ggsave(filename = file.path(results_path, "silhouette_plot.png"),
  plot     = sil_plot,
  width    = 200,       # cm (1/3 of 84.1 cm)
  height   = 150,     # cm (aspect ratio 10:6)
  units    = "mm",
  dpi      = 300)
```

```{r}
# Fit the clustering on selected K
set.seed(123)

esm_kmed_cent <- pam(centroid_dep[, c("Geom_1", "Geom_2", "Geom_3", "Geom_4", "Geom_5")], k = 4, metric = "euclidean", stand = FALSE)

centroid_dep$cluster_med <- esm_kmed_cent$cluster

centroid_dep <- centroid_dep |>
  ungroup() |>
  dplyr::mutate(cluster_med = forcats::as_factor(cluster_med),
                cluster_med = forcats::fct_relevel(cluster_med, "1"),
                GENDER = forcats::as_factor(GENDER),
                AGE_stnd = scale(AGE)
  ) 
```

```{r}
table(esm_kmed_cent$cluster)
```
```{r}
# Bootstrap cluster and stability measures
set.seed(1)

cb <- clusterboot(centroid_dep[, c("Geom_1", "Geom_2", "Geom_3", "Geom_4", "Geom_5")],
                  B = 1000,
                  clustermethod = pamkCBI,
                  k = 4,
                  count = FALSE)

cb$bootmean          # Jaccard per cluster
cb$bootbrd           # times each cluster "dissolved"
cb$bootrecover
```
```{r}
comp_labels <- c(
  Geom_1 = "Negative Intrusive \n thinking",
  Geom_2 = "Goal-Directed \nthinking",
  Geom_3 = "Mental Time \nTravel",
  Geom_4 = "Positive \nEngagement",
  Geom_5 = "Modality"
)

medoid_profiles <- as.data.frame(esm_kmed_cent$medoids) |>
  mutate(cluster = factor(row_number())) |>
  tidyr::pivot_longer(-cluster, names_to = "component", values_to = "value") |>
  mutate(component = factor(comp_labels[component], levels = comp_labels))

hp_clusters <-ggplot(medoid_profiles, aes(component, cluster, fill = value)) +
  geom_tile(colour = "white", linewidth = 0.5) +
  geom_text(aes(label = round(value, 2)), size = 5) +
  scale_fill_gradient2(low = "#4575b4", mid = "white", high = "#d73027",
                       midpoint = 0, name = "Score (SD)") +
  labs(x = "Default thought pattern dimension", y = "Cluster") +
  theme_minimal() +
  theme(panel.grid = element_blank(),
        axis.text.x  = element_text(size = 10),
        axis.text.y  = element_text(size = 10),
        axis.title.y = element_text(margin = margin(r = 10), size = 12),
        axis.title.x = element_text(margin = margin(t = 10), size = 12))

hp_clusters
```
```{r}
ggsave(filename = file.path(results_path, "heatmap_clusters.png"),
  plot     = hp_clusters ,
  width    = 200,       # cm (1/3 of 84.1 cm)
  height   = 150,     # cm (aspect ratio 10:6)
  units    = "mm",
  dpi      = 300)
```

## Cluster depression relation
```{r}
# Fit a model to test if the cluster membership predicts levels of depression
cluster_centroid <- lm(CESD_stnd ~ cluster_med + AGE_stnd + GENDER, data = centroid_dep)

p <- parameters::parameters(cluster_centroid) |> as.data.frame()
cl <- grepl("cluster_med", p$Parameter)
p$p_fdr <- NA
p$p_fdr[cl] <- p.adjust(p$p[cl], method = "fdr")
p
```

```{r}
# Effect size
sd_radius <- sd(centroid_dep$CESD_stnd, na.rm = TRUE)

parameters::parameters(cluster_centroid) |>
  as.data.frame() |>
  dplyr::filter(grepl("cluster_med", Parameter)) |>
  dplyr::mutate(d       = Coefficient / sd_radius,
                d_low   = CI_low  / sd_radius,
                d_high  = CI_high / sd_radius)
```

```{r}
emm_cl <- estimate_means(cluster_centroid, by = "cluster_med") |> as.data.frame()

emm_cl

p_dep <- ggplot(emm_cl, aes(Mean, cluster_med)) +
  geom_pointrange(aes(xmin = CI_low, xmax = CI_high)) +
  labs(y ="Cluster membership", x = "Depression symptoms \n(estimated marginal mean)") +
  theme_minimal()
p_dep
```
# Distance analysis

```{r}
radius_mixed <- lmer(Radius_geom ~ Stimulus_Type + AGE_stnd + GENDER +  (Stimulus_Type| ID), data = esm_centroids)

parameters::parameters(radius_mixed, p_adjust = "fdr")
```

```{r}
# Effect sizes
sd_radius <- sd(esm_centroids$Radius_geom, na.rm = TRUE)

parameters::parameters(radius_mixed) |>
  as.data.frame() |>
  dplyr::filter(grepl("Stimulus_Type", Parameter)) |>
  dplyr::mutate(d       = Coefficient / sd_radius,
                d_low   = CI_low  / sd_radius,
                d_high  = CI_high / sd_radius)
```

```{r}
esm_stim_cent <- esm_centroids |>
  mutate(Condition = recode(Stimulus_Type,
                            "Baseline" = "Baseline",
                            "pos" = "Positive",
                            "neg" = "Negative")
  )

cols <- c("Positive" = "#cc0000", "Negative" = "#0c60ac", "Baseline" = "#444444")

violin_distance <- ggplot(esm_stim_cent, aes(x = Condition, y = Radius_geom, fill = Condition)) +
  geom_violin(position = position_dodge(width = 0.9), 
              width = 0.8, 
              alpha = 0.6, 
              show.legend = FALSE) +
  geom_boxplot(position = position_dodge(width = 0.9), 
               width = 0.1, color = "#444444", 
               alpha = 0.2, 
               show.legend = FALSE) +
  scale_fill_manual(values = cols) +
  ylim(0, 5) +
  labs(x = "Condition", y = "Distance from centroid") +
  theme_minimal(base_size = 13) +
  theme(
    axis.line = element_line(linewidth = 0.8, colour = "black", linetype = 1),
    panel.grid.minor = element_blank(),
    axis.title.x = element_text(margin = margin(t = 10)),
    axis.title.y = element_text(margin = margin(r = 10))
  )

ggsave(
  filename = file.path(results_path, "violin_distance.svg"),
  plot     = violin_distance,   
  units    = "mm",
  dpi      = 300
)
```

## Leave one out cross validation
```{r}
# leave one out, compute geom, the calculate distance between left out observation and new geom
loo_df <- esm_traits_base |>
  group_by(ID) |>
  group_modify(~ {
    M <- as.matrix(dplyr::select(.x, Component_1_z:Component_5_z))
    tibble(Radius_geom = sapply(seq_len(nrow(M)), function(i) {
      gm <- geo_median(M[-i, , drop = FALSE])$p
      sqrt(sum((M[i, ] - gm)^2))
    }))
  }) |>
  ungroup()
```

```{r}
# Join new baseline distances with original emotional distances from default
loo_df$Stimulus_Type <- "Baseline"

esm_radius_emo <- esm_centroids |>
  filter(Stimulus_Type != "Baseline") |>
  dplyr::select(ID, Stimulus_Type, Radius_geom)

loo_df <- loo_df |>
  rbind(esm_radius_emo)
```

```{r}
loo_df <- loo_df |>
  left_join(cesd_by_id, by = "ID")
```

```{r}
radius_mixed_loo <- lmer(Radius_geom ~ Stimulus_Type + AGE_stnd + GENDER + (Stimulus_Type| ID), data = loo_df)

parameters::parameters(radius_mixed_loo, p_adjust = "fdr")
```
```{r}
# Effect sizes
sd_radius <- sd(loo_df$Radius_geom, na.rm = TRUE)

parameters::parameters(radius_mixed_loo) |>
  as.data.frame() |>
  dplyr::filter(grepl("Stimulus_Type", Parameter)) |>
  dplyr::mutate(d       = Coefficient / sd_radius,
                d_low   = CI_low  / sd_radius,
                d_high  = CI_high / sd_radius)
```

```{r}
cols <- c("pos" = "#cc0000", "neg" = "#0c60ac", "Baseline" = "#444444")

violin_distance_loo <- ggplot(loo_df, 
                              aes(x = Stimulus_Type, y = Radius_geom, fill = Stimulus_Type)) +
  geom_violin(position = position_dodge(width = 0.9), 
              width = 0.8, 
              alpha = 0.6, 
              show.legend = FALSE) +
  geom_boxplot(position = position_dodge(width = 0.9), 
               width = 0.1, 
               color = "#444444", 
               alpha = 0.2, 
               show.legend = FALSE) +
  scale_fill_manual(values = cols) +
  scale_x_discrete(labels = c("Baseline", "Negative","Positive")) +
  ylim(0, 5) +
  labs(x = "Condition", y = "Distance from leave-one-out centroid") +
  theme_minimal(base_size = 13) +
  theme(
    axis.line = element_line(linewidth = 0.8, colour = "black", linetype = 1),
    panel.grid.minor = element_blank(),
    axis.title.x = element_text(margin = margin(t = 10)),
    axis.title.y = element_text(margin = margin(r = 10))
  )

violin_distance_loo

ggsave(
  filename = file.path(results_path, "violin_distance_loo.svg"),
  plot     = violin_distance_loo,   
  units    = "mm",
  dpi      = 300
)
```

# Angle analysis

```{r}
# Directional displacement analysis
centroid_neg <- esm_centroids |>
  filter(Stimulus_Type == "neg") 

centroid_pos <- esm_centroids |>
  filter(Stimulus_Type == "pos")

centroid_pos <- centroid_pos |>
  group_by(ID) |>
  summarise(across(Component_1_z:Component_5_z, \(x) mean(x, na.rm = TRUE)), .groups = "drop")

centroid_neg <- centroid_neg |>
  group_by(ID) |>
  summarise(across(Component_1_z:Component_5_z, \(x) mean(x, na.rm = TRUE)), .groups = "drop")

component_names <- paste0("Component_", 1:5, "_z")
cent_names <- paste0("Geom_", 1:5)
```
```{r}
identical(centroid_neg$ID, centroid_summary$ID)
identical(centroid_pos$ID, centroid_neg$ID)
```

```{r}
# Dot product of vectors a * b = ||a||*||b||cos0
delta_neg <- centroid_neg[, component_names] - centroid_summary[, paste0(cent_names)]
delta_pos <- centroid_pos[, component_names] - centroid_summary[, paste0(cent_names)]
dot <- rowSums(delta_pos*delta_neg)
```

```{r}
# Vector norms
norm_neg <- sqrt(rowSums(delta_neg^2))
norm_pos <- sqrt(rowSums(delta_pos^2))
```

```{r}
# Cosine of angle (clip to avoid invalid arccos)
cos_theta <- dot / (norm_neg * norm_pos)
cos_theta <- pmin(pmax(cos_theta, -1), 1)  # Clamp to [-1, 1]
```
```{r}
# Angle in degrees
angles_deg <- acos(cos_theta) * (180 / pi)
```

```{r}
# Combine into a data frame
angle_df <- data.frame(
  ID = centroid_neg$ID,
  Angle_Pos_Neg = angles_deg
)

mean_delta_neg <- colMeans(delta_neg)
mean_delta_pos <- colMeans(delta_pos)

group_dot  <- sum(mean_delta_neg * mean_delta_pos)
group_norm <- sqrt(sum(mean_delta_neg^2)) * sqrt(sum(mean_delta_pos^2))
group_angle <- acos(group_dot / group_norm) * (180 / pi)

t_test <- t.test(angle_df$Angle_Pos_Neg, mu = 0)
cat("Group-level angle (Pos vs. Neg):", round(group_angle, 2), "degrees\n")
cat("Mean individual angle:", round(mean(angle_df$Angle_Pos_Neg), 2), "degrees\n")
cat("t =", round(t_test$statistic, 3), 
    ", p =", format.pval(t_test$p.value, digits = 3), 
    ", N =", nrow(angle_df), "\n")
```
### Permutation of angle direction

```{r}
# Normalise positive and negative vectors, calculate the difference between them, and measure the length of this difference, then permute the positive and negative labels 10000 times and repeat.
range(norm_pos); range(norm_neg)      # nothing near zero

U_pos <- as.matrix(delta_pos) / norm_pos
U_neg <- as.matrix(delta_neg) / norm_neg

D   <- U_pos - U_neg
obs <- sqrt(sum(colMeans(D)^2))

set.seed(1)
null <- replicate(10000, {
  s <- sample(c(1, -1), nrow(D), replace = TRUE)
  sqrt(sum(colMeans(D * s)^2))
})

p_perm <- (sum(null >= obs) + 1) / (length(null) + 1)
c(observed = obs, p = p_perm)
```
```{r}
quantile(null, c(.5, .95, .99))
round(colMeans(D), 3)
```
```{r}
perm_angle <- ggplot(data.frame(null = null), aes(null)) +
  geom_histogram(bins = 60, fill = "grey80", colour = "white") +
  geom_vline(xintercept = obs, colour = "#d73027", linewidth = 0.8) +
  annotate("text", x = obs, y = Inf, label = "Observed", hjust = 1.1, vjust = 2, size = 3) +
  labs(x = "Length of mean difference vector\n (Positive - Negative vector)",
       y = "Count") +
  theme_minimal()

perm_angle
```
```{r}
ggsave(
  filename = file.path(results_path,"perm_angle.svg"),
  plot     = perm_angle,   
  units    = "mm",
  dpi      = 300
)
```

### Components driving the divergence

```{r}
# Difference between pos and neg deltas
delta_diff <- delta_pos[, 1:5] - delta_neg[, 1:5]
delta_diff <- delta_diff[complete.cases(delta_diff), ]  # remove any NAs
```
```{r}
# Compute mean, SD, and t-tests for each component
means <- colMeans(delta_diff)
sds   <- apply(delta_diff, 2, sd)
t_res <- map_dfr(delta_diff, ~ tidy(t.test(.x, mu = 0)), .id = "Component")
cohen_ds <- map_dfr(delta_diff, ~ as.data.frame(effectsize::cohens_d(.x, mu = 0)), .id = "Component")
```
```{r}
# Assemble into summary table
divergence_by_component <- tibble(
  Component  = names(means),
  Mean_Diff  = means,
  Std_Diff   = sds
) |>
  left_join(t_res |> dplyr::select(Component, statistic,conf.low, conf.high, parameter, p.value), by = "Component") |>
  left_join(cohen_ds |> dplyr::select(Component, Cohens_d), by = "Component") |>
  rename(t_value = statistic, p_value = p.value) |>
  arrange(p_value) 
```

```{r}
# View table
divergence_by_component |> mutate(across(where(is.numeric), ~round(., 3)),
                                  p_fdr = p.adjust(p_value, method = "fdr")) 
```

```{r}
#Step 1: Compute means of displacement vectors
mean_delta_pos <- colMeans(delta_pos[, 1:5])
mean_delta_neg <- colMeans(delta_neg[, 1:5])
```

```{r}
# Step 2: Create human-readable labels
labels <- c(
  "Component_1_z" = "Negative Intrusive Thought",
  "Component_2_z" = "Goal Directed Thinking",
  "Component_3_z" = "Mental Time Travel",
  "Component_4_z" = "Positive Engagement",
  "Component_5_z" = "Modality"
)
```

```{r}
# Step 3: Assemble into summary table
direction_summary <- tibble(
  Component = names(mean_delta_pos),
  Label     = labels[names(mean_delta_pos)],
  Mean_Pos_Displacement = mean_delta_pos,
  Mean_Neg_Displacement = mean_delta_neg
)
```

```{r}
# View table
print(direction_summary)

direction_radar <- direction_summary |>
  dplyr::select(Label, Mean_Pos_Displacement, Mean_Neg_Displacement) |>
  pivot_longer(cols = starts_with("Mean_"), names_to = "Condition", values_to = "Value") |>
  pivot_wider(names_from = Label, values_from = Value)
```

```{r}
# Create the max and min rows as one-row data frames
max_row <- as.data.frame(matrix(1, nrow = 1, ncol = ncol(direction_radar)))
min_row <- as.data.frame(matrix(-1, nrow = 1, ncol = ncol(direction_radar)))
```

```{r}
# Copy column names from your main data
colnames(max_row) <- colnames(direction_radar)
max_row$Condition <- "Max"
colnames(min_row) <- colnames(direction_radar)
min_row$Condition <- "Min"
```

```{r}
# Bind all together
radar_dir <- bind_rows(max_row, min_row, direction_radar)
png(file.path(results_path,"radar_plot_displacement.png"), width = 1200, height = 1000, res = 300)
```

```{r}
# Expand plotting area: reduce margins
op <- par(mar = c(1, 1, 1, 1))  # bottom, left, top, right
```

```{r}
# Set colors and labels
colors_border <- c("#cc0000", "#0c60ac")  # Positive, Negative
labels_order <- radar_dir$Condition
```

```{r}
# Plot
fmsb::radarchart(radar_dir|> dplyr::select(-Condition),
           axistype = 1,
           seg = 4,
           pcol = colors_border,
           plwd = 2,
           plty = 1,
           cglcol = "grey", cglty = 1,
           axislabcol = "grey", caxislabels = c(-1, -0.5, 0, 0.5, 1), cglwd = 1,
           vlcex = 0.8)
legend(x = "bottomright", title = "Condition", legend = c("Positive", "Negative"),
       bty = "n", pch = 20, col = colors_border, text.col = "black", cex = 1, pt.cex = 2)
# Reset plotting parameters
par(op)
```

# Mood Model

## Component 1

```{r}
mood_depression_mdl <- lmerTest::lmer(Component_1_z ~ Stimulus_Type * CESD_stnd + AGE_stnd + GENDER  + (Stimulus_Type|ID), data = esm_traits_z)

ci_dep_int_param <- parameters::parameters(mood_depression_mdl)
ci_dep_int_param |> knitr::kable(digits = 3)
```

```{r}
# Effect sizes for main effects that are not factor contrasts
effectsize::eta_squared(mood_depression_mdl, partial = TRUE, alternative = "two.sided")
```

```{r}
cols <- c("Baseline" = "#595959", "pos" = "#cc0000", "neg" = "#0c60ac")

pred <- estimate_relation(mood_depression_mdl)

depression_model_plot <- pred |> 
  ggplot(aes(x=CESD_stnd, y=Predicted)) +
  geom_line(aes(color=Stimulus_Type), linewidth =0.8) + 
  geom_ribbon(aes(fill=Stimulus_Type, ymin=CI_low, ymax=CI_high), alpha=0.1) +
  geom_point(data = esm_traits,
             aes(x = CESD_stnd, y = Component_1, color = Stimulus_Type),
             alpha = 0.25, size = 1.2, inherit.aes = FALSE, position = "jitter") +
  labs( y="Negative Intrusive Thought scores", x="Depressive symtpom score (CESD)", fill="Condition") +
  scale_color_manual(aesthetics = c("colour", "fill"), labels = c("Daily-life", "Positive", "Negative"), values = cols) +
  guides(color=guide_legend("Condition")) +
  theme_minimal(base_size = 20) +
    theme(
      panel.grid.major     = element_blank(),
      panel.grid.minor     = element_blank(),
      axis.line            = element_line(color = "black", linewidth = 0.7),
      axis.ticks           = element_line(color = "black"),
      # inside top-right   
      legend.position      = "top",
      legend.location = "panel",
      legend.key.size      = unit(0.7, "cm"),
      legend.text          = element_text(size = 18),
      legend.title = element_text(face = "bold"),
    )

depression_model_plot
```

```{r}
ggsave(file.path(results_path, "depression_int_c1.svg"),
  plot     = depression_model_plot,
  width    = 200,       # cm (1/3 of 84.1 cm)
  height   = 155,     # cm (aspect ratio 10:6)
  units    = "mm",
  dpi      = 300
)
```

```{r}
ec <- estimate_contrasts(mood_depression_mdl, contrast = "Stimulus_Type", by="CESD_stnd", p_adjust = "fdr", length=3) 

ec |> knitr::kable(digits = 4)
```

```{r}
# Effect size for factor contrasts
sd_y <- sd(esm_traits_z$Component_1_z, na.rm = TRUE)

ec |>
  mutate(d      = Difference / sd_y,
         d_low  = CI_low  / sd_y,
         d_high = CI_high / sd_y)
```
# PCA robustness checks

## Cat PCA

```{r}
# Fit categorical PCA
items <- esm_traits_base |> dplyr::select(c(Task, Future:Words))
items[] <- lapply(items, function(x) ordered(x, levels = 1:5))

cat5 <- princals(items,
                 ndim    = 5,
                 ordinal = TRUE,
                 degrees = -1,
                 knots   = knotsGifi(items, type = "D"))
```


```{r}
tr <- cat5$transform
trlist <- if (is.list(tr) && !is.data.frame(tr)) {
  lapply(tr, function(x) as.numeric(as.matrix(x)[, 1]))
} else {
  setNames(lapply(colnames(tr), function(v) as.numeric(tr[, v])), colnames(tr))
}

stopifnot(all(names(items) %in% names(trlist)))
stopifnot(length(trlist[[1]]) == nrow(items))

quant <- do.call(rbind, lapply(names(items), function(v) {
  d <- unique(data.frame(
    item  = v,
    cat   = as.integer(as.character(items[[v]])),
    quant = trlist[[v]]
  ))
  d[order(d$cat), ]
}))
```
```{r}
lin <- t(sapply(split(quant, quant$item), function(d) {
  d <- d[order(d$cat), ]
  s <- diff(d$quant)
  c(r          = cor(d$cat, d$quant),
    min_step   = min(s),
    step_ratio = max(s) / min(s),
    n_distinct = length(unique(round(d$quant, 6))))
}))
round(lin, 3)
```

```{r}
# Rotate
L_cat <- cat5$loadings
rot   <- stats::varimax(L_cat)     # normalize = TRUE by default
L_cat_rot <- unclass(rot$loadings)
```

```{r}
# Check eigenvalues
cat5$evals[1:18]
```
### Compare loadings
```{r}
# Load PCA loadings from original solution
pca_loadings  <- read.csv(file.path(dirname(getwd()), "data", "pca_loadings_with_task.csv"), row.names = 1) 

pca_loadings_og <- as.matrix(pca_loadings[, -1])
rownames(pca_loadings_og) <- pca_loadings[[1]]

# Create matrix of cat PCA loadings
cat_pca_loadings <- as.matrix(L_cat_rot)
```

```{r}
# COmpute congruence between two loading sets
C <- psych::factor.congruence(pca_loadings_og , cat_pca_loadings)
C |> knitr::kable(digits = 3)
```

```{r}
# Cosine similarities between the two sets of loadings
cosines <- svd(t(qr.Q(qr(pca_loadings_og))) %*% qr.Q(qr(cat_pca_loadings)))$d
round(cosines, 3)
```
## Similarity analysis of mean centered PCA

```{r}
# Import person mean centered PCA scores from python output
data_path_cent <- file.path(dirname(getwd()), "data", "pca_scores_centred.csv")

if (!file.exists(data_path_cent)) {
  stop("Dataset is currently not publicaly available.")
}
```

```{r}
pca_centered <- read.csv(data_path_cent) |>
  dplyr::mutate(ID = forcats::as_factor(ID)) 
```

```{r}
# Filter out ppts that are not in esm_trait base 
ids_keep <- unique(esm_traits_base$ID)

pca_centered <- pca_centered |>
  filter(ID %in% ids_keep) |>
  mutate(ID = droplevels(ID))

length(levels(pca_centered$ID))
```
```{r}
pca_centered <- pca_centered |>
  rename("Component_1" = "Component.1",
         "Component_2" = "Component.2",
         "Component_3" = "Component.3",
         "Component_4" = "Component.4",
         "Component_5" = "Component.5")
```

```{r}
participant_ids_cent <- unique(pca_centered$ID)
```

```{r}
# Create an empty data frame to store results
within_similarity_cent_df <- data.frame(ID = character(), within_similarity_cent_z = numeric(), stringsAsFactors = FALSE)
```

```{r}
# Loop over each participant ID
for (id in participant_ids) {
  # Subset data for the current participant
  participant_data <- pca_centered |>
    dplyr::filter(ID == id) |>
    dplyr::select(c("Component_1":"Component_5"))
  # Convert to matrix
  participant_matrix <- as.matrix(participant_data)
  # Only proceed if participant has more than 1 measurement
  if (nrow(participant_matrix) > 1) {
    # Compute trial × trial correlation
    trial_cor <- cor(t(participant_matrix))
    # Extract lower triangle
    lower_vals <- trial_cor[lower.tri(trial_cor)]
    #Fisher transform correlations before averaging
    lower_vals_clipped <- pmin(pmax(lower_vals, -0.9999), 0.9999)
    z_vals <- atanh(lower_vals_clipped)
    # Compute mean z-value
    mean_z <- mean(z_vals, na.rm = TRUE)
    # Add result to the data frame
    within_similarity_cent_df <- rbind(within_similarity_cent_df, data.frame(ID = id, within_similarity_cent_z = mean_z))
  }
}
mean_within_similarity_cent <- mean(within_similarity_cent_df$within_similarity_cent_z)
```

```{r}
# Between subject similarity
mean_vectors_cent <- pca_centered |>
  group_by(ID) |>
  summarise(across(c("Component_1":"Component_5"), \(x) mean(x, na.rm = TRUE))) |>
  ungroup()
mean_matrix_cent <- as.matrix(mean_vectors_cent[, -1])
```

```{r}
# Correlation matrix 
between_cor_cent <- cor(t(mean_matrix_cent))
```

```{r}
# extract lower triangle
lower_between_cent <- between_cor_cent[lower.tri(between_cor_cent)]
```

```{r}
# Fisher z
z_between_cent <- atanh(lower_between_cent)
mean_between_similarity_cent <- mean(z_between_cent, na.rm = TRUE)
within_z_cent <- within_similarity_cent_df$within_similarity_cent_z
between_z_cent <- z_between_cent  # the full lower triangle of between-subject correlations, Fisher-z transformed
```

```{r}
# Compute the observed difference
observed_diff_cent <- mean(within_z_cent) - mean(between_z_cent)
```

```{r}
# Combine all values into one vector
all_z_values_cent <- c(within_z_cent, between_z_cent)
group_labels <- c(rep("within", length(within_z_cent)), rep("between", length(between_z_cent)))
```

```{r}
# Set number of permutations
n_perm <- 10000
perm_diffs_cent <- numeric(n_perm)
set.seed(123)
for (i in 1:n_perm) {
  shuffled_labels <- sample(group_labels)
  group1 <- all_z_values_cent[shuffled_labels == "within"]
  group2 <- all_z_values_cent[shuffled_labels == "between"]
  perm_diffs_cent[i] <- mean(group1) - mean(group2)
}
```

```{r}
# p-value: how often is permuted diff ≥ observed diff?
p_value_cent <- mean(perm_diffs_cent >= observed_diff_cent)
perm_cent_df <- data.frame(perm_diff_cent = perm_diffs_cent)
```
```{r}
observed_diff_cent
p_value_cent
```

```{r}
# Make the plot
permutation_plot <- ggplot(perm_cent_df, aes(x = perm_diff_cent)) +
  geom_histogram(bins = 50, fill = "grey80", color = "black") +
  geom_vline(xintercept = observed_diff_cent, color = "#990000", linewidth = 1.2) +
  annotate("text",
           x = Inf,
           y = Inf,
           label = "Observed\nDifference",
           hjust = 0.99, vjust = 1, color = "#990000", size = 5) +
  scale_x_continuous(limits = range(c(perm_diffs_cent, observed_diff_cent))) +
  scale_y_continuous(limits = range(0,1000)) +
  labs(
    x = "Mean thought similarity difference (within - between)",
    y = "Frequency"
  ) +
  theme_minimal(base_size = 20) +
  theme(
    plot.margin = unit(c(1, 2, 1, 2), "cm"),
    panel.grid.major     = element_blank(),
    panel.grid.minor     = element_blank(),
    axis.line            = element_line(color = "black"),
    axis.ticks           = element_line(color = "black"),
    # inside top-right
    legend.position      = c(0.997, 0.998),
    legend.justification = c(1, 1),
    legend.key.size      = unit(0.4, "cm"),
    legend.text          = element_text(size = 18),
    legend.spacing.y = unit(0,"cm"),
  ) 
ggsave(
   filename = file.path(results_path, "permutation_plot_centered_pca.svg"),
   plot     = permutation_plot,
   width    = 200,       # cm (1/3 of 84.1 cm)
   height   = 150,     # cm (aspect ratio 10:6)
   units    = "mm",
   dpi      = 300
 )
permutation_plot
```

# ICC

```{r}
esm_dimensions <- esm_traits_base |> 
  dplyr::select(c("ID", "Task_stnd","Future_stnd":"Words_stnd"))
```
```{r}
items <- esm_dimensions |>
  dplyr::select(-c("ID")) |> 
  colnames() 
```

```{r}
# across all items
sapply(items, function(it) {
  icc(lmer(reformulate("(1|ID)", response = it), data = esm_dimensions))$ICC_adjusted
}) |> knitr::kable(digits = 3)
```
```{r}
# ICC for multilevel data
comps <- paste0("Component_", 1:5, "_z")
icc_comp <- sapply(comps, function(v) {
  m <- lmer(reformulate("(1|ID)", response = v), data = esm_traits_base)
  vc <- as.data.frame(VarCorr(m))$vcov
  vc[1] / sum(vc)
})
round(icc_comp, 3)
```
# PCA space stability

```{r}
esm_traits_base <- esm_traits_base |>
  group_by(ID) |>
  arrange(Date, .by_group = TRUE) |>
  mutate(obs_n = row_number()) |>
  ungroup()

icc_by_obs <- map_dfr(3:36, function(k) {
  d <- filter(esm_traits_base, obs_n <= k)
  tibble(k = k,
         n_participants = n_distinct(d$ID),
         component = paste0("Component_", 1:5, "_z"),
         icc = sapply(paste0("Component_", 1:5, "_z"), function(v) {
           m <- lmer(reformulate("(1|ID)", response = v), data = d)
           vc <- as.data.frame(VarCorr(m))$vcov
           vc[1] / sum(vc)
         }),
         median_obs = d |> count(ID) |> pull(n) |> median())
})

sd_by_obs <- map_dfr(3:36, function(k) {
  ids_ok <- esm_traits_base |> count(ID) |> filter(n >= k) |> pull(ID)
  esm_traits_base |>
    filter(ID %in% ids_ok, obs_n <= k) |>
    group_by(ID) |>
    summarise(across(Component_1_z:Component_5_z, ~ sd(.x, na.rm = TRUE)), .groups = "drop") |>
    summarise(across(starts_with("Component"), ~ mean(.x, na.rm = TRUE))) |>
    mutate(k = k, n_participants = length(ids_ok))
})
```

```{r}
ggplot(icc_by_obs, aes(k, icc)) +
  geom_line() +
  geom_point(aes(size = median_obs)) +
  facet_wrap(~ component, ncol = 3) +
  scale_size_continuous(range = c(0.5, 2.5)) +
  labs(x = "Observations per participant (k)",
       y = "ICC (proportion of variance between participants)",
       size = "Meadian number of observations provided by participants") +
  theme_minimal()
```

```{r}
sd_long <- sd_by_obs |>
  tidyr::pivot_longer(starts_with("Component"),
                      names_to = "component", values_to = "sd")

tp_stability <-ggplot(sd_long, aes(k, sd)) +
  geom_line() + geom_point() +
  facet_wrap(~ component, ncol = 3) +
  labs(x = "Observations per participant (k)",
       y = "Mean within-participant SD") +
  theme_minimal()

tp_stability
```

```{r}
icc_by_obs |>
  group_by(component) |>
  arrange(k) |>
  mutate(icc_final = icc[k == max(k)],
         reached = icc >= 0.99 * icc_final) |>
  filter(reached) |>
  slice(1) |>
  dplyr::select(component, k_stable = k)
```

```{r}
sd_long |>
  group_by(component) |>
  arrange(k) |>
  mutate(plateau = mean(sd[k >= 12 & k <= 20]),
         prop    = (sd - first(sd)) / (plateau - first(sd))) |>
  filter(prop >= 0.95) |>
  slice(1) |>
  dplyr::select(component, k_stable = k)
```

# Effect of dataset

```{r}
esm_traits_base <- esm_traits_base |>
  mutate(Dataset = forcats::as_factor(Dataset))

esm_centroids <- esm_centroids |>
  mutate(Dataset = forcats::as_factor(Dataset))
```

```{r}
radius_mixed_check <- lmerTest::lmer(Radius_geom ~ Stimulus_Type * Dataset + AGE_stnd + GENDER + (Stimulus_Type | ID),
                                     data = esm_centroids)

anova(radius_mixed_check, type = 3) |> knitr::kable(digits = 3)
```
```{r}
emmeans::emmeans(radius_mixed_check, ~ Dataset | Stimulus_Type)
```

```{r}
mood_mixed_check <- lmerTest::lmer(Component_1_z ~ Stimulus_Type*Dataset + AGE_stnd + GENDER + (Stimulus_Type| ID), data = esm_traits_z)

anova(mood_mixed_check, type = 3) |> knitr::kable(digits = 3)
```
## Effect of depression and dataset

```{r}
esm_traits_base <- esm_traits_base |>
  group_by(Dataset) |>
  mutate(CESD_within = CESD_stnd - mean(CESD_stnd, na.rm = TRUE),
         CESD_between = mean(CESD_stnd, na.rm = TRUE)) |>
  ungroup()
  
```

```{r}
c1_dep_check <- lme4::lmer(Component_1_z ~ CESD_within + CESD_between + AGE_stnd + GENDER + (1 | ID), data = esm_traits_base)

parameters::parameters(c1_dep_check)
```
```{r}
esm_binomial <- esm_binomial|>
  left_join(esm_traits_base |> dplyr::select(ID, Dataset) |> distinct(), by = "ID") |>
  group_by(Dataset) |>
  mutate(CESD_within = CESD_stnd - mean(CESD_stnd, na.rm = TRUE),
         CESD_between = mean(CESD_stnd, na.rm = TRUE)) |>
  ungroup()
```

```{r}
model_binom_check <- glm(cbind(high, total - high) ~ CESD_within + CESD_between + AGE_stnd + GENDER ,
                   family = quasibinomial(link = "logit"),
                   data = esm_binomial)
summary(model_binom_check)
or_bi <- parameters::parameters(model_binom_check, exponentiate = TRUE) 
or_bi |> knitr::kable(digits = 3)
```

```{r}
centroid_dep_check <- centroid_dep |>
  group_by(Dataset) |>
  mutate(CESD_within = CESD_stnd - mean(CESD_stnd, na.rm = TRUE),
         CESD_between = mean(CESD_stnd, na.rm = TRUE)) |>
  ungroup()
```

```{r}
# Fit a model to test if the cluster membership based on the centroid coordinates predicts levels of depression
cluster_centroid_check <- lm(CESD_within ~ cluster_med + AGE_stnd + GENDER , data = centroid_dep_check)

parameters::parameters(cluster_centroid_check)
```
```{r}
# Fit a model to test if the cluster membership based on the centroid coordinates predicts levels of depression
cluster_centroid_check <- lm(CESD_between ~ cluster_med + AGE_stnd + GENDER , data = centroid_dep_check)

parameters::parameters(cluster_centroid_check)
```

```{r}
esm_traits_z <- esm_traits_z |>
  group_by(Dataset) |>
  mutate(CESD_within = CESD_stnd - mean(CESD_stnd, na.rm = TRUE),
         CESD_between = mean(CESD_stnd, na.rm = TRUE)) |>
  ungroup()

c1_dep_check <- lme4::lmer(Component_1_z ~ Stimulus_Type * CESD_within + CESD_between +
                 AGE_stnd + GENDER + (Stimulus_Type | ID), data = esm_traits_z)

parameters::parameters(c1_dep_check)
```
